---
title: "Bayesian Methods"
date: "April 2025"
author: "Mark R. Williamson"
output: html_document
---
# Part 1: Setup

## References
[Bayesian One-Sample T-Test (Stan)](https://cran.r-project.org/web/packages/bridgesampling/vignettes/bridgesampling_stan_ttest.html)
[Bayes Factor](https://cran.r-project.org/web/packages/BayesFactor/vignettes/manual.html)

https://learningstatisticswithr.com/book/bayes.html#bayesregression
https://michael-franke.github.io/Bayesian-Regression/practice-sheets/01b-regression-BRMS.html
https://www.bayesrulesbook.com/chapter-8.html
https://search.r-project.org/CRAN/refmans/bayesreg/html/bayesreg-package.html

## Get Libraries
```{r library, include=FALSE}
library(bridgesampling)
#library(rstan)
library(BayesFactor)
library(ggplot2)
library(sdamr)
library(MASS)
library(lme4)
library(rpart)
library(dplyr)
library(survival)
library(ggfortify)
library(BayesSurvival)
library(bayesreg)
library(corrplot)

#https://www.r-inla.org/

#library(INLA)
#library(parallel)
```
***

## Data
```{r datasets}

#One-sample t-test
data(sleep)

#Two-sample t-test
  #chickwts

#Paired t-test
sleep_diff <- sleep$extra[sleep$group == 1] - sleep$extra[sleep$group == 2] #group 1-2

#Non-parametric t-test
data(chickwts)
## Restrict to two groups
chickwts = chickwts[chickwts$feed %in% c("horsebean","linseed"),]
## Drop unused factor levels
chickwts$feed = factor(chickwts$feed)
## Plot data
plot(weight ~  feed, data = chickwts, main = "Chick weights")

#1-Way ANOVA
data(ToothGrowth)
ToothGrowth$dose = factor(ToothGrowth$dose)
levels(ToothGrowth$dose) = c("Low", "Medium", "High")

#2-Way ANOVA
  #ToothGrowth

#Repeated Measures ANOVA
data(puzzles)

#Correlation
data(iris)

#Simple Linear Regression
data(attitude)

#Multiple Linear Regression
  #attitude

#Logistic Regression
data(spambase)
spambase$is.spam <- factor(spambase$is.spam)
spambase2<-sample_n(spambase, 500)

#Poisson Regression
data(warpbreaks)

#Proportion Test
  #coin flip

#Chi-Square Test of Independence
data(survey)
table(survey$W.Hnd)
table(survey$Sex)
Shand_df <- table(survey$W.Hnd, survey$Sex)

#Generalized Linear Mixed Model
data(cake)

#Survival Analysis
data(veteran)
veteran$trt <- as.factor(veteran$trt)
levels(veteran$trt) <- c("standard", "test")
veteran$prior <- as.factor(veteran$prior)
levels(veteran$prior) <- c("No", "Yes")

#BEYOND

#Structural Equation Model

#Generalized Estimating Equation

#Generalized Additive Model
```
***
# Part 2: T-Tests

## One-Sample T-test
```{r one-sample-t-test}
#Graph
sleep$x <- c(rep('x',20))
ggplot(data=sleep, aes(x=x, y=extra)) +
  geom_boxplot(width=0.25)+
  geom_dotplot(binaxis='y', stackdir='center')+
  geom_hline(yintercept=0, color="red", linewidth=1)+
  labs(y="Extra Sleep", x="")

#Traditional
t_ttest1 <- t.test(x=sleep$extra, alternative="two.sided", mu=0, paired=FALSE)
t_ttest1

#Bayesian
b_ttest1 <-ttestBF(x=sleep$extra, mu=0)
b_ttest1 #strong evidence for alternative hyp.

b_ttest1_chains <- posterior(b_ttest1, iterations = 10000)
summary(b_ttest1_chains)
plot(b_ttest1_chains[,1:2])

```
***

## Two-Sample T-test
```{r two-sample-t-test}
#Graph
ggplot(data=chickwts, aes(x=feed, y=weight, fill=feed)) +
  geom_boxplot()+
  geom_dotplot(binaxis='y', stackdir='center')

#Traditional
t_ttest2 <- t.test(weight ~ feed, data = chickwts, var.eq=TRUE)
t_ttest2

#Bayesian
b_ttest2 <-ttestBF(formula=weight~feed, data=chickwts)
b_ttest2 #moderate evidence for alternative hyp.

b_ttest2_chains <- posterior(b_ttest2, iterations = 10000)
summary(b_ttest2_chains)
plot(b_ttest2_chains[,2]) #horsebean- linseed

```
***

## Paired t-test
```{r paired-t-test}

#Graph
ggplot(data=sleep, aes(x=group, y=extra, group=ID)) +
  geom_line() + 
  labs(y="Extra Sleep")


#Traditional
t_ttest3a <- t.test(x=sleep$extra[sleep$group == 1], y=sleep$extra[sleep$group == 2],
                    alternative="two.sided", paired=TRUE)
t_ttest3a

t_ttest3b <- t.test(x=sleep_diff, alternative="two.sided", mu=0, paired=FALSE)
t_ttest3b

#Bayesian
b_ttest3 <-ttestBF(x=sleep_diff)
b_ttest3 #strong evidence for alternative hyp.

b_ttest3_chains <- posterior(b_ttest3, iterations = 10000)
summary(b_ttest3_chains)
plot(b_ttest3_chains[,1:2])

```
***
# Part 3: ANOVA

## One-Way ANOVA
```{r one-way-ANOVA}
#issue ref: https://github.com/richarddmorey/BayesFactor/issues/169
#ggplot ref: https://mspeekenbrink.github.io/sdam-r-companion/bayesian-hypothesis-testing-with-bayes-factors.html

#Graph
ggplot(data=ToothGrowth, aes(x=dose, y=len, fill=dose)) +
  geom_boxplot()

#Traditional
t_anova1 <-aov(len ~ dose, data=ToothGrowth)
summary(t_anova1)

em_dose <- emmeans(t_anova1, ~ dose)
contrast(em_dose, "pairwise", adjust = "Tukey")

#Bayesian
b_anova1 <-anovaBF(len ~ dose, data=ToothGrowth)
b_anova1  #extreme evidence for alternative hyp.

b_anova1_chains <- posterior(b_anova1, iterations=10000)
summary(b_anova1_chains)
plot(b_anova1_chains[,2:4])

colMeans(b_anova1_chains) #get names of columns

b_anova1_plot <-data.frame(category=c(rep("Low",10000), rep("Medium",10000), rep("High",10000)),
                  values=c(b_anova1_chains[,c("dose-Low")], b_anova1_chains[,c("dose-Medium")],
                           b_anova1_chains[,c("dose-High")]))

ggplot(data=b_anova1_plot, aes(x=values, fill=category))+
  geom_density(alpha=0.25)

```
***

## Two-Way ANOVA
```{r two-way-ANOVA}
#Graph
ggplot(data=ToothGrowth, aes(x=dose, y=len, fill=supp)) +
  geom_boxplot()

#Traditional
t_anova2 <-aov(len ~ dose + supp + dose:supp, data=ToothGrowth)
summary(t_anova2)

em_dose_supp <- emmeans(t_anova2, ~ dose:supp)
contrast(em_dose_supp, "pairwise", adjust = "Tukey")

#Bayesian
b_anova2 <-anovaBF(len ~ dose + supp + dose:supp, data=ToothGrowth)
b_anova2  #large effect of dose, small effect of supplement, maybe an interaction
  #Index [4] is full model

#check top models
anovaBF(len ~ supp*dose, data=ToothGrowth, whichModels="top") #removing any effect reduces model fit
  
b_anova2_chains <- posterior(b_anova2, index=4, iterations=10000)
summary(b_anova2_chains[,1:13])
plot(b_anova2_chains[,7:12])

colMeans(b_anova2_chains) #get names of columns

#WRONG - NIX FOR NOW
b_anova2_plot <-data.frame(
                  category=c(rep("OJ-LOW",10000), rep("OJ-Med",10000), rep("OJ-High",10000),
                             rep("VC-LOW",10000), rep("VC-Med",10000), rep("VC-High",10000)
                             ),
                  values=c(b_anova2_chains[,c("supp:dose-OJ.&.Low")], b_anova2_chains[,c("supp:dose-OJ.&.Medium")],
                           b_anova2_chains[,c("supp:dose-OJ.&.High")], b_anova2_chains[,c("supp:dose-VC.&.Low")],
                           b_anova2_chains[,c("supp:dose-VC.&.Medium")], b_anova2_chains[,c("supp:dose-VC.&.High")]
                           ))

ggplot(data=b_anova2_plot, aes(x=values, fill=category))+
  geom_density(alpha=0.25)
```
***

## Repeated-Measures ANOVA
```{r rm-ANOVA}
#Graph
ggplot(data=puzzles, aes(x=shape, y=RT, fill=color)) +
  geom_boxplot() +
  scale_fill_manual(values=c("orange","white"))

#Traditional
t_anova3 <-aov(RT ~ shape*color + Error(ID/(shape*color)), data=puzzles)
summary(t_anova3)

em_shape_color <- emmeans(t_anova3, ~ shape:color)
contrast(em_shape_color, "pairwise", adjust = "Tukey")

#Bayesian
b_anova3 <- anovaBF(RT ~ shape*color + ID, data = puzzles, 
             whichRandom="ID") 
b_anova3 # Main effects model has strongest support [index=3]
plot(b_anova3)

b_anova3_chains <- posterior(b_anova3, index=3, iterations=10000)
plot(b_anova3_chains[,2:3])
plot(b_anova3_chains[,4:5])

  #compare random effects to non-random effects
bfWithoutID <- lmBF(RT ~ shape*color, data = puzzles)
bfWithoutID

bfOnlyID <- lmBF(RT ~ ID, whichRandom="ID",data = puzzles)
bfOnlyID

bf2 <- bfWithoutID / bfOnlyID
bf2

b_anova3[4] / bf2 #random effects much preferred
  
b_anova3_chains <- posterior(b_anova3, index=3, iterations=10000)
summary(b_anova3_chains)

colMeans(b_anova3_chains) #get names of columns

#PLOTS NOT WORKING
b_anova3_plot <-data.frame(
                  Shape=c(rep("round",10000), rep("square",10000)),
                  Color=c(rep("color",10000), rep("monoch",10000)),
                  S_values=c(b_anova3_chains[,c("shape-round")], b_anova3_chains[,c("shape-square")]),
                  C_values=c(b_anova3_chains[,c("color-color")], b_anova3_chains[,c("color-monochromatic")]))
                  
ggplot(data=b_anova3_plot, aes(x=S_values, fill=Shape))+
  geom_density(alpha=0.25)
ggplot(data=b_anova3_plot, aes(x=C_values, fill=Color))+
  geom_density(alpha=0.25)

b_anova3_chains2 <- posterior(b_anova3, index=4, iterations=10000)
colMeans(b_anova3_chains2)

b_anova3_plot2 <-data.frame(
                  category=c(rep("round+color",10000), 
                             rep("round+monoch",10000), 
                             rep("square+color",10000),
                             rep("square+monoch",10000)
                             ),
                  values=c(b_anova3_chains2[,c("shape:color-round.&.color")],
                           b_anova3_chains2[,c("shape:color-round.&.monochromatic")],
                           b_anova3_chains2[,c("shape:color-square.&.color")],
                           b_anova3_chains2[,c("shape:color-square.&.monochromatic")]
                           ))

ggplot(data=b_anova3_plot2, aes(x=values, fill=category))+
  geom_density(alpha=0.25)

b_anova3_plot3 <-data.frame(
                  category=c(rep("round",10000), 
                             rep("square",10000), 
                             rep("color",10000),
                             rep("monoch",10000)
                             ),
                  values=c(b_anova3_chains2[,c("shape-round")],
                           b_anova3_chains2[,c("shape-square")],
                           b_anova3_chains2[,c("color-color")],
                           b_anova3_chains2[,c("color-monochromatic")]
                           ))

ggplot(data=b_anova3_plot3, aes(x=values, fill=category))+
  geom_density(alpha=0.25)


```
***
# Part 4: Correlation and Regression

#Correlation
```{r correlation}
#Graph
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width))+
  geom_point()+
  geom_smooth(method="lm", se=FALSE) 

#Traditional
t_cor1 <- cor.test(y = iris$Sepal.Length, x = iris$Sepal.Width)
t_cor1
  #cor = -0.118

#Bayesian
b_cor1 <- correlationBF(y = iris$Sepal.Length, x = iris$Sepal.Width)
b_cor1 #Anecdotal evidence for null hypothesis

b_cor1_chains <- posterior(b_cor1, iterations=10000)
summary(b_cor1_chains)
plot(b_cor1_chains[,"rho"])
```
***

#Simple Linear Regression
```{r simple-linear-regression}
#Graph
ggplot(data=attitude, aes(x=raises, y=rating))+
  geom_point()+
  geom_smooth(method="lm", se=TRUE) 
    # Estimate = 0.69
    # Intercept = 19.98

#Traditional
t_reg1 <-lm(rating ~ raises, data=attitude)
summary(t_reg1)

#Bayesian
b_reg1 <-regressionBF(rating ~ raises, data=attitude)
b_reg1 #very strong evidence for alternative hypothesis

b_reg1_chains <- posterior(b_reg1, index=1, iterations=10000)
summary(b_reg1_chains)
  #Estimate = 0.62
```
***

#Multiple Linear Regression
```{r multiple-linear regression}
#Graph
corrplot(cor(attitude))

#Traditional
t_reg2 <-lm(rating ~ raises + complaints + privileges + learning + critical + advance, data=attitude)
summary(t_reg2) #complaints significant

#Bayesian
b_reg2 <-regressionBF(rating ~ raises + complaints + privileges + learning + critical + advance, data=attitude)
length(b_reg2) #63 models

#full model
b_reg2["raises + complaints + privileges + learning + critical + advance"]

#top 10 models
head(b_reg2, n=10) #complaints only is far and away the best

#compare top models
b_reg2_comp <- head(b_reg2) / max(b_reg2)
b_reg2_comp 
plot(b_reg2_comp)

#examining trad vs. bayesian model outputs
t_reg2B <- lm(rating ~ complaints, data=attitude)
summary(t_reg2B)

b_reg2B <- regressionBF(rating ~ complaints, data=attitude)
b_reg2_chains <- posterior(b_reg2B, index=1, iterations=10000)
summary(b_reg2_chains)
```
***

#Logistic Regression
```{r logistic-regression}

#Graph
  #N/A

#Traditional
t_logistic <- glm(is.spam ~ word.freq.money + word.freq.free + 
                            word.freq.credit+ word.freq.font +
                            word.freq.edu   + word.freq.1999,
                            data=spambase2, family="binomial")
summary(t_logistic)

#Bayesian
b_logistic <- bayesreg(is.spam ~ word.freq.money + word.freq.free + 
                                 word.freq.credit+ word.freq.font +
                                 word.freq.edu   + word.freq.1999,
                                 data=spambase2, model="logistic")
summary(b_logistic)
```
***

#Poisson Regression
```{r poisson-regression}

#Graph
ggplot(data=warpbreaks, aes(y=breaks, x=wool)) + 
  geom_boxplot()
ggplot(data=warpbreaks, aes(y=breaks, x=tension)) + 
  geom_boxplot()
ggplot(data=warpbreaks, aes(y=breaks, x=wool, fill=tension)) + 
  geom_boxplot()

#Traditional
t_poisson <- glm(breaks ~ wool*tension, data=warpbreaks, family="poisson")
summary(t_poisson)

em_wool_tension <- emmeans(t_poisson, ~ wool:tension)
contrast(em_wool_tension, "pairwise", adjust = "Tukey")

#Bayesian
b_poisson <- bayesreg(breaks ~ wool*tension, data=warpbreaks, model="poisson",
                      burnin=10000)
summary(b_poisson)
```
***
# Part 5: Frequency Analysis

#Proportion Test
```{r proportion}
#Graph
  #N/A

#Traditional
t_prop <- binom.test(x=63, n=100, p=0.5)
t_prop #significantly different from null

#Bayesian
b_prop <- proportionBF(y=63, N=100, p=0.5)
b_prop #moderate evidence for alternative

b_prop_chains <- posterior(b_prop, iterations=10000)
plot(b_prop_chains[,"p"])
```
***

#Chi-Square Test of Independence
```{r chi-square-ind}
#Graph

#Traditional
Shand_df[1]/Shand_df[2] *100 #percentage of left-handed females = 6.4%
Shand_df[3]/Shand_df[4] *100 #percentage of left-handed males   = 9.3%

t_freq <- chisq.test(Shand_df) 
t_freq #not significant

#Bayesian
b_freq <- contingencyTableBF(Shand_df, sampleType = "indepMulti", fixedMargin = "cols")
b_freq #moderate evidence for the null

b_freq_chains <- posterior(b_freq, iterations = 10000)
plot(b_freq_chains)

LeftGivenMale  <- b_freq_chains[,"pi[1,1]"] / b_freq_chains[,"pi[*,1]"]
LeftGivenFemale<- b_freq_chains[,"pi[1,2]"] / b_freq_chains[,"pi[*,2]"]

plot(mcmc(LeftGivenMale  - LeftGivenFemale))

```
***
# Part 6: Advanced Models

#Generalized Linear Mixed Model
```{r}

#Graph
ggplot(data=cake, aes(x=angle))+
  geom_histogram(bins=12, fill='grey', color='black')

ggplot(data=cake, aes(y=angle, x=recipe, fill=recipe))+
  geom_boxplot()
ggplot(data=cake, aes(y=angle, x=temperature, fill=temperature))+
  geom_boxplot()

#Traditional
t_glmm0 <- lm(angle ~ recipe + temperature, data=cake)
summary(t_glmm0)

t_glmm1 <- lmer(angle ~ recipe + temperature + (1|replicate), data=cake)
summary(t_glmm1)

t_glmm2 <- glmer(angle ~ recipe + temperature + (1|replicate), family=poisson, data=cake)
summary(t_glmm2)

t_glmm3 <- glmer.nb(angle ~ recipe + temperature + (1|replicate), data=cake)
summary(t_glmm3)


#Bayesian
b_glmm <- generalTestBF(angle ~ recipe + temperature + replicate, data=cake, whichRandom="replicate")
b_glmm #best model is temperature + replicate

b_glmm_chains <- posterior(b_glmm, index=6, iterations=10000)
summary(b_glmm_chains)
plot(b_glmm_chains[,1:2])

Temp_175 <- b_glmm_chains[,"temperature-175"]
Temp_225 <- b_glmm_chains[,"temperature-225"]

plot(mcmc(Temp_225-Temp_175))

#Additional
Temp_215 <- b_glmm_chains[,"temperature-215"]

plot(mcmc(Temp_225-Temp_215))
plot(mcmc(Temp_215-Temp_175))
```
***

#Survival Analysis
```{r}

#ref: https://becarioprecario.bitbucket.io/inla-gitbook/ch-survival.html#non-parametric-estimation-of-the-survival-curve

#data formatting
veteran$time.m <- round(veteran$time / 30, 3)

#Traditional
t_surv <- Surv(veteran$time.m, veteran$status)

t_km_1 <- survfit(t_surv ~ 1)
t_km_2 <- survfit(t_surv ~ -1 + trt, data = veteran)

t_cox <- coxph(t_surv ~ trt, data=veteran)
summary(t_cox)

autoplot(t_km_1)
autoplot(t_km_2)

#Bayesian
b_surv_1 <- BayesSurv(df=veteran, time="time.m", event="status")

PlotBayesSurv(b_surv_1, object = "survival")
PlotBayesSurv(b_surv_1, object = "cumhaz")
PlotBayesSurv(b_surv_1, object = "hazard")

plot(b_surv_1$surv.post.mean, b_surv_1$surv.eval.grid)


#models for each treatment, then plot
table(veteran$trt)

veteran_stnd <-veteran[veteran$trt=="standard",]
veteran_test <-veteran[veteran$trt=="test",]

b_surv_s <- BayesSurv(df=veteran_stnd, time="time.m", event="status")
b_surv_t <- BayesSurv(df=veteran_test, time="time.m", event="status")

b_surv_trt_plot_ds <- data.frame(Treatment=c(rep("standard",50), rep("test",50)),
                                 Time     =c(b_surv_s$surv.eval.grid, b_surv_t$surv.eval.grid),
                                 Survival =c(b_surv_s$surv.post.mean, b_surv_t$surv.post.mean))
head(b_surv_trt_plot_ds)

ggplot(data=b_surv_trt_plot_ds, aes(x=Time, y=Survival, group=Treatment, colour=Treatment))+
  geom_line(linewidth=2)
  #no confidence intervals, but okay

```
***

#Structural Equation Model
```{r}

#JAGS format
#https://www.jstatsoft.org/article/view/v085i04
  #BEYOND SCOPE

```
***

#Generalized Estimating Equation
```{r}

#BEYOND SCOPE
```
***

#Generalized Additive Model
```{r}
#STAN format
#https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.14498
  #BEYOND SCOPE


```
***
#Non-parametric models
```{r}
#https://link.springer.com/chapter/10.1007/978-3-319-59650-1_24
```

